Mapping out the punishment strategy space

### 2024-07-24 appendix.Rmd
submit manuscript

\setcounter{page}{1}
\centering
\noindent \hspace*{10mm} \small Why do people punish? Evidence for a range of strategic concerns \newline
\hspace*{1cm} \small Scott Claessens\textsuperscript{1}, Quentin D. Atkinson\textsuperscript{1}, & Nichola Raihani\textsuperscript{1,2} \newline
\raggedright
\noindent \footnotesize \textsuperscript{1} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{2} Department of Experimental Psychology, University College London, London, United Kingdom \newline
\par
\noindent{\bfseries{Contents}}
\contentsline {subsection}{Supplementary Figures S1-S13}{2}{}
\contentsline {subsection}{Supplementary Tables S1-S4}{15}{}
\normalsize
\newpage

@@

@@

```{r tableSurveyWordings}
apa_table(
  tar_read(tableSurveyWordings),
  landscape = TRUE,
  longtable = TRUE,
  align = c("p{6cm}", "p{11cm}", "p{3cm}"),
  caption = "(ref:tableSurveyWordingsCaption)",
  booktabs = TRUE,
  placement = "H",
  fontsize = "scriptsize"
)
```

>>>>

\setcounter{page}{1}
\centering
\noindent \hspace*{25mm} \small Why do people punish? Evidence for a range of strategic concerns \newline
\hspace*{30mm} \small Scott Claessens\textsuperscript{1}, Quentin D. Atkinson\textsuperscript{1}, & Nichola Raihani\textsuperscript{1,2} \newline
\raggedright
\noindent \footnotesize \textsuperscript{1} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{2} Department of Experimental Psychology, University College London, London, United Kingdom \newline
\par
\noindent{\bfseries{Contents}}
\contentsline {subsection}{Supplementary Figures S1-S14}{2}{}
\contentsline {subsection}{Supplementary Tables S1-S4}{16}{}
\normalsize
\newpage

@@

(ref:plotSplineCaption) *The relationships between strategy frequencies and the
overall costs of strategies across all twelve punishment decisions, in both
countries.* Each point is a unique strategy that appears in our dataset at
least once (for ease of presentation, the "never punish" strategy is excluded).
Lines and shaded areas represent posterior predictions from splines fitted to
the trend in each country separately.

```{r plotSpline, fig.width=6, fig.height=3.5, fig.cap="(ref:plotSplineCaption)", warning=FALSE}
tar_read(plotSpline)
```

@@

```{r tableSurveyWordings}
apa_table(
  tar_read(tableSurveyWordings),
  landscape = TRUE,
  longtable = TRUE,
  align = c("p{6cm}", "p{11cm}", "p{3cm}"),
  caption = "(ref:tableSurveyWordingsCaption)",
  booktabs = TRUE,
  placement = "H",
  font_size = "scriptsize"
)
```

### 2024-07-24 manuscript.Rmd
submit manuscript

bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This manuscript has not yet been peer-reviewed. \par This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 \text{license:} \url{https://doi.org/10.31234/osf.io/ys6rm} \par \text{Classification:} Social Sciences; Psychological and Cognitive Sciences \par \text{Keywords:} punishment; cooperation; economic games}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

@@

# Significance Statement

Research has shown that humans are willing to punish others at a personal cost,
but the motives for this behaviour remain unclear. Using a suite of economic 
games, we show that punitive behaviour is consistent with a variety of strategic
concerns. While most people choose not to punish in our games, people who do 
punish tend to be motivated by egalitarianism rather than deterrence or 
norm enforcement. We also find that different punishment strategies are linked 
with different character traits, including personality, prosocial preferences, 
and political and religious views. This research highlights the multifaceted
nature of human punishment.

\newpage
\linenumbers

@@

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment strategies.
Table \@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness [@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives [but see [@Jordan2020]].

@@

It is striking that the most common strategy in our dataset was to never punish. 
This is partly because punishment in these games imposes an economic cost for no 
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper 
to punish, we may have seen more punishment from participants. Indeed, 
`r printnum(mean(tar_read(dExc)$SelfRate10[tar_read(dExc)$strategy == "Never punish"] > 0, na.rm = TRUE)*100, digits = 0)`% 
of participants following the never punish strategy positively stated that they 
didn't want to pay to reduce anyone's bonus but would have done so if it were 
free. But the frequency of the never punish strategy perhaps also reflects a 
more general aversion to peer punishment, an aversion that has been highlighted 
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples
[@Balafoutas2014;@Balafoutas2016] and in small-scale societies [@Baumard2010]. 
One reason that people may be averse to peer punishment is that, because it is 
a fundamentally harmful act, punishment can reflect badly on the punisher and 
people might therefore refrain from punishing others to avoid reputational 
damage [@Raihani2015]. People frequently avoid taking actions that could harm 
their reputation, even when they don’t know if reputation is at stake (as in the 
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation [@Raihani2019]. This may be
especially likely in situations that lack clear institutional norms to 
legitimise punishment, such as our economic games. As with reputation damage, 
people might abstain from peer punishment to avoid retaliation, regardless of 
whether retaliation is actually possible. By contrast, institutionalised 
punishment in small-scale societies often functions to compensate victims while 
limiting the potential for feuds and cycles of retaliation
[@Fitouchi2023;@Singh2022]. Future research should uncover whether people are 
more willing to punish in these conventionalised contexts [e.g., see
[@Molleman2019]].

We finally highlight some key limitations that can guide future research. One 
limitation with our design is that some strategies required more punishment than 
others to be met. For example, the competitive strategy required punishment in 
ten of twelve decisions, compared to the avoid disadvantageous inequity strategy
which required only one instance of punishment (Table \@ref(tab:tableStrategies)). 
Strategies thus differed in how "expensive" they were to implement, perhaps 
explaining why some strategies were more common than others. This issue is 
largely unavoidable in our design since strategies, by their very nature, differ 
in how punitive they are. To partially mitigate this issue, we employed the 
strategy method to incentivise participants, such that payoffs were calculated 
from a randomly chosen game instead of summed across all games.

Another limitation is that our results may be contingent on the particular suite 
of anonymous stealing games that we used. With our anonymous design, we were 
unable to study other potential reputational strategies underlying punishment, 
such as signalling trustworthiness [@Jordan2020]. Moreover, stealing may be 
evaluated differently to other forms of cheating, such as not contributing to 
public goods, and other negative behaviours, such as lying or breaking taboos. 
Finally, we were unable to include all permutations of situational features in 
the games (e.g., second-party vs. third-party, equal vs. unequal) making it 
difficult to interpret some patterns of behaviour. For example, many 
participants punished only in the third-party game (Game F), but it is not clear 
whether these participants were driven by the third-party nature of the game or 
simply by the fact that stealing in that game generated inequality. Future work 
should determine whether different reputational contexts, target behaviours, and 
combinations of situational features elicit different punishment strategies.

@@

\begingroup

<div id="refs" custom-style="Bibliography"></div>
\endgroup

>>>>

bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This manuscript has not yet been peer-reviewed. \par This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 \text{license:} \url{https://doi.org/10.31234/osf.io/ys6rm} \par \text{Classification:} Social Sciences; Psychological and Cognitive Sciences \par \text{Keywords:} punishment; cooperation; economic games}
appendix          : "appendix.Rmd"
csl               : nature.csl
notes-after-punctuation: false
---

@@

\linenumbers

@@

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment strategies.
Table \@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness [@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives (but see ref[@Jordan2020]).

@@

It is striking that the most common strategy in our dataset was to never punish. 
This is partly because punishment in these games imposes an economic cost for no 
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper 
to punish, we may have seen more punishment from participants. Indeed, 
`r printnum(mean(tar_read(dExc)$SelfRate10[tar_read(dExc)$strategy == "Never punish"] > 0, na.rm = TRUE)*100, digits = 0)`% 
of participants following the never punish strategy positively stated that they 
didn't want to pay to reduce anyone's bonus but would have done so if it were 
free. But the frequency of the never punish strategy perhaps also reflects a 
more general aversion to peer punishment, an aversion that has been highlighted 
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples
[@Balafoutas2014;@Balafoutas2016] and in small-scale societies [@Baumard2010]. 
One reason that people may be averse to peer punishment is that, because it is 
a fundamentally harmful act, punishment can reflect badly on the punisher and 
people might therefore refrain from punishing others to avoid reputational 
damage [@Raihani2015]. People frequently avoid taking actions that could harm 
their reputation, even when they don’t know if reputation is at stake (as in the 
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation [@Raihani2019]. This may be
especially likely in situations that lack clear institutional norms to 
legitimise punishment, such as our economic games. As with reputation damage, 
people might abstain from peer punishment to avoid retaliation, regardless of 
whether retaliation is actually possible. By contrast, institutionalised 
punishment in small-scale societies often functions to compensate victims while 
limiting the potential for feuds and cycles of retaliation
[@Fitouchi2023;@Singh2022]. Future research should uncover whether people are 
more willing to punish in these conventionalised contexts (e.g., see ref
[@Molleman2019]).

One potential limitation with our design is that some strategies required more 
punishment than others, meaning that some strategies were more "expensive" to 
implement. For example, the competitive strategy required punishment in ten of 
twelve decisions, compared to the avoid disadvantageous inequity strategy which 
required only one instance of punishment (Table \@ref(tab:tableStrategies)).
We employed the strategy method to deal with this, calculating participant 
payoffs from a randomly chosen game instead of accumulating the costs across 
all games. This mitigates the concern that the total cost is the most salient
feature determining participants' decisions. But even accounting for this, 
one could still argue that the difference in overall costs explains why, for 
example, the competitive strategy is less common in our dataset than the avoid 
disadvantageous inequity strategy. We do not think this feature of our design 
is a concern, however, for a number of reasons. First, we are interested in 
measuring strategies underlying *costly* punishment. Some of these strategies,
by their very nature, will be more costly to implement than others. This is 
reflected in our design. It would not make theoretical sense to use an 
alternative design where more or less punitive strategies are manipulated to 
cost participants the same amount. Second, when we asked participants whether
they would have punished if it were free to do so, we found low agreement 
with this statement (Supplementary Figure \@ref(fig:plotSliders1)), 
suggesting that participants were not particularly sensitive to the costs of 
punishment. In line with this, when we plot the frequencies of different 
strategies against their expected costs, we find that cost is not a perfect
predictor of strategy frequency: many "cheap" strategies are rare in our 
data and some "expensive" strategies are quite common (e.g., always punish;
Supplementary Figure \@ref(fig:plotSpline)). Third, this argument implies 
that the high frequency of the never punish strategy is merely an artefact
of our design, since it is the only strategy that does not cost anything to
implement. But as we have discussed, many other studies have found similar
aversions to costly punishment in the lab [@Barclay2016;@Batistoni2022;
@Bone2015;@Deutchman2021;@Raihani2012b] and in the real world
[@Balafoutas2014;@Balafoutas2016;@Baumard2010] suggesting that this result is
not an artefact of our design.

@@

\begingroup
\setlength\parindent{0pt}
<div id="refs" custom-style="Bibliography"></div>
\endgroup
